\documentclass[onecard,frame]{flashcards}

\cardfrontstyle[\large\slshape]{headings}
\cardbackstyle{empty}
\cardfrontfoot{Category Theory}

%\usepackage{amsfonts}
\usepackage{amssymb,amsmath,tikz-cd,xspace}
\usepackage[T1]{fontenc}

\begin{document}

\input{tex.macros}

\long\def\Comdiag#1{
\begin{center}
\begin{tikzcd}[ampersand replacement=\&]
#1
\end{tikzcd}
\end{center}
}

\Card[Intro]{Author, copyleft}{Ben Klemens, CC BY/SA/NC}

\Card{Domain of $f$}{ For $f:X\to Y$, $X$.}
\Card{Codomain of $f$}{ For $f:X\to Y$, $Y$.}

\Card{Composition of $f$ and $g$}{
As a diagram,
\Comdiag{
X \arrow[r,"f"] \&Y \arrow[r, "g"] \& Z
}

Written $g \circ f: X\to Z$.
}

\Card{$\Hom{set}(X,Y)$}{The set of functions $X\to Y$}

\Card{Equality of $f$ and $g$}{
    $f:X\to Y$ and $g:X\to Y$ are equal iff $f(x)=g(x), \forall x \in X$
}

\Card{Identity function on $X$}{
    ${\rm id}_X:X\to X$ such that ${\rm id}_X(x) = x, \forall x\in X$.
}

\Card{Isomorphism}{
A one-to-one correspondence.

\Comdiag{
    X \arrow[r,"\cong"] \&Y
}
There exists a function $g:Y\to X$ such that
$g \circ f = {\rm id}_X$ and
$f \circ g = {\rm id}_Y$.
}

\Card[Characteristics]{Properties of an isomorphism}{
\items{
\item Reflexive: Any set is isomorphic to itself.
\item Commutative: if $f:A\to B$ has inverse $g:B\to A$, then $g$ has inverse $f$.
\item Transitive: $A$ isomorphic to $B$ and $B$ isomorphic to $C$ $\Rightarrow$ $A$ is
isomorphic to $C$.
}
}

\card{Sets are isomorphic}{There exists at least one isomorhpism between them.}

\Card[Notation]{$\underline n$}{The set $\{1, 2,\dots n\}$}

\Card{Cardinality of finite set $X$}{$|X| = n$ iff

\Comdiag{
    X \arrow[r,"\cong"] \&\underline n
}
}

\Card[Lemma]{Cardinality of $A$ and $B$  given
    \Comdiag{A \arrow[r,"\cong"] \&B}}{|A| = |B|}

\Card{Injective}{
    For $F:A\to B$, $f(x) = f(y)$ only if $x=y$: points map to different values.
}


\Card{A diagram commutes}{
    All paths from top left to bottom right are equal. E.g.,
    \Comdiag{
    A \ar[r, "f"] \ar[dr,"h"'] \& B\ar[d, "g"]\\
            \& C
    }
    commutes iff $g\circ f = h$.
}

\card{Endomap}{
    A map whose domain and codomain are the same, so of the form $f:A\to A$.
}

\card{Lifting problem}{
    aka {\em choice}
    What are the functions $f$ that allow this diagram to commute?
    \Comdiag{
        X \ar[r, dashrightarrow, "f"] \ar[dr, "h"] \& Y \arrow [d, "g"]\\
        \& Z
    }
}

\card{Determination problem}{
    aka {\em extension}
    What are the functions $g$ that allow this diagram to commute?
    \Comdiag{
        X \ar[r, "f"] \ar[dr, "h"] \& Y \arrow [d, dashrightarrow, "g"]\\
        \& Z
    }
    If any, $h$ is determined by $f$.
}
\card{monotone transformation}{If $a \leq b$ then $f(a) \leq f(b)$}




\Card{Product of two sets}{
    The set of ordered pairs. For sets $X$ and $Y$,
    $$\{(x,y)|x\in X, y\in Y\}.$$
}

\Card{Projection function $\pi_i$}{
    Given a product $X\times Y$ with elements $(x, y)$, $\pi_1: X\times Y \to X$ produces
    $x$. Similarly for $\pi_2$.

    \Comdiag{
        \& X \times Y \ar[dl,"\pi_1"'] \ar[dr,"\pi_2"]\\
        X \&  \& Y 
    }
}

\Card{Universal property of projections}{
    For any set A and any $f:A\to X$ and $g:A\to Y$, $\exists$ a unique $f\times g$ so this commutes:
    \Comdiag{
        \& X\times Y \ar[dl,"\pi_1"'] \ar[dr,"\pi_2"] \\
    X \& \& Y \\
    \& A \ar[ul, "\forall f"] \ar[ur, "\forall g"'] \ar[uu, dashrightarrow, "f\times g"]
    }

}

\Card[Notation]{$X\sqcup Y$}{
    The union of the elements of $X$ and $Y$, via indicator functions $i_1: X\to X\sqcup Y$ and $i_2$:

    \Comdiag{
        X \ar[dr,"i_1"'] \& \& Y \ar[dl,"i_2"] \\
        \& X\sqcup Y
    }
    Elements of $X\sqcup Y$ have a source indicator, so if $z\in X$ and $z\in Y$,
    $i_1z, i_2z \in X\sqcup Y$.
}

\Card{Universal property of coproducts}{
    For any $f:X\to A$, $g:Y\to A$, $\exists$ a unique $f\sqcup g:X\sqcup Y\to A$:
    \Comdiag{
        \& A \\
        X \ar[dr,"i_1"'] \ar[ur,"\forall f"] \& \& Y \ar[dl,"i_2"] \ar[ul,"\forall g"']\\
        \& X\sqcup Y \ar[uu, dashrightarrow, "f\sqcup g"]
    }
}

\Card{Fiber product}{
    Given \Comdiag{
        \& Y \ar[d,"g"]\\
        X \ar[r, "f"] \& Z
    }
    $X \times_Z Y \equiv \{(x,w,y)| f(x)=w=g(y)\}$.
}

\Card{Pullback}{
    Any set $W$ isomorphic to the fiber product $X\times_Z Y$, giving the commutative diagram
    \Comdiag{
        W \ar[r, "\pi_2"] \ar[d, "\pi_1"'] \& Y \ar[d,"g"]\\
        X \ar[r, "f"] \& Z
    }
}

\Card{Universal property for pullbacks}{
    Given $t:X\to Z$ and $u:Y\to Z$,
    $\exists$ a unique function $f\times_Z g: A\to X\times_z Y$ such that
    \Comdiag{
        X\times_Z Y \arrow[drr, bend left, "\pi_1"] \arrow[ddr, bend right, "\pi_2"] \& \& \\
        \& A\arrow[ul, dotted, "f\times_Z g" description] \arrow[r, "\forall f"] \arrow[d, "\forall g"]
        \& Y \arrow[d, "u"] \\
        \& X \arrow[r, "t"] \& Z
    }
}

\Card{Coequalizer}{
    Given the equivalence relation $\{(f(x), g(x))|x\in X\}\subseteq Y\times Y$,
$Coeq(f,g)\equiv Y \backslash f(x)\sim g(x)$
    \Comdiag{
    Z \arrow[r, shift left,"f"] \arrow[r,"g"'] \&Y \arrow[r] \& Coeq(f,g)
    }
}

\Card{Initial set}{
    A set $S$ such that for every set $A$ there exists a unique mapping $S\to A$.
}

\Card{Retract section, retract projection}{
    Given $f:X\to Y$, $g:Y\to X$ and $g\circ f =id_X$,
    \items{
        \item $f$ is a retract section
        \item $g$ is a retract projection
    }
}

\Card[Notation]{$Y^X$}{$Y^X\equiv \Hom{Set}(X, Y)$.}

\Card[Proposition]{Currying}{
    For any sets $A, X, Y$, $\exists$ a bijection
$$C: \Hom{Set}(X\times A, Y) \to \Hom{Set}(X, \Hom{Set}(A, Y))$$
i.e.,
$$C: Y^{X\times A} \to Y^{A^X}$$
}

\Card{Span}{
    \Comdiag{
         \& R \ar[dl, "f"'] \ar[dr, "g"]\\
        X\&   \& Y
    }

}

\Card{Composite Span}{
    Two spans joined by their fiber product.

    \Comdiag{
        \& \& R\times_Y R' \ar[dl] \ar[dr]\\
         \& R \ar[dl, "f"'] \ar[dr, "g"]\& \& R' \ar[dl, "f'"'] \ar[dr, "g'"]\\
        X\&   \& Y \& \& Z
    }

}

\Card{Equalizer}{
$Eq(f, g)\equiv {x\in X|f(x)=f(y)}$.

    \Comdiag{Eq(f, g) \ar[r] \& X \ar[r, shift left, "f"] \ar[r, "g"'] \& Y
    }
}

\Card{Terminal set}{
    A set $S$ such that for every set $X$, there exists a unique mapping $f:X\to S$.
}

\Card{Limit}{
    The unique mapping to sets (e.g., $X$ and $Y$) that allows a given cone diagram to
    commute (see the universal property cards), with a unique mapping to corresponding
    mappings after transformations of $X$ and $Y$ ($f(x)$, $g(y)$). 
}

\Card[Notation]{$x\sim_R y$ or $x\sim y$}{
    The equivalence relation, $R\subseteq X\times X$, where
    \items{
        \item $(x,x)\in R$ (reflexive)
        \item $(x,y)\in R \Leftrightarrow (y, x)\in R$ (symmetry)
        \item $(x,y)\in R + (y, z)\in R \rightarrow (y, z)\in R$ (transitivity)
    }
    $x\sim_R y \equiv (x,y) \in R.$
}

\Card{Equivalence class}{
    Given an equivalence relation $\sim$ on the set $X$, a set $A$ where
    \items{
        \item $A\neq \emptyset$
        \item $x\in A$ and $y\in A  \rightarrow x\sim y$
        \item $x\in A$ and $x\sim y  \rightarrow y\in A$
    }
}

\Card{Quotient $X\backslash \sim$}{The set of equivalence classes of $\sim$ in $X$.}

\Card{Fiber sum}{
    A {\em pushout}: given $f:Z\to X$ and  $g:Z\to Y$, the fiber sum $X\sqcup_Z Y\equiv (X\sqcup Z \sqcup
Y)\backslash \sim$ where $z\sim f(z)$ and $z\sim g(z)$. So,

    \Comdiag{
        Z \ar[d, "f"'] \ar[r, "g"] \& Y \ar[d, "i_2"] \\
        X \ar[r, "i_1"] \& X\sqcup_Z Y
    }
}

\Card{Universal property for pushouts}{
    \small{Given $t:Z\to X$ and $u:Z\to Y$,
    $\exists$ a unique function $f\sqcup_Z g:  X\sqcup_z Y \to A$ such that
    \Comdiag{
        Z \ar[d, "t"'] \ar[r, "u"] \& Y \ar[d, "g"] \ar[ddr, bend left, "i_2"] \\
        X \ar[r, "f"] \ar[drr, bend right, "i_1"] \& A \\
    \& \& X\sqcup_Z Y \arrow[ul, dotted, "f\sqcup_Z g" description]
    }
    So, $f = (f\sqcup_Z g)\circ i_1$ and $g = (f\sqcup_Z g)\circ i_2$.}
}

\Card{Power set}{For a given set $S$, the set of subsets of $S$. Here notated ${\mathbb P}(S)$}

\card{Notation}{$\Omega$}{The set $\{$True, False $\}$}

\Card[Proposition]{For an arbitrary set $S$, how do $\Omega$ and ${\mathbb P}(S)$ relate?}{
$\exists$ an isomporphism $f:$\Comdiag{ \Hom{Set}(S,\Omega) \arrow[r,"\cong"] \& {\mathbb P}(S) }
}

\Card{Characteristic function}{For a set $S'\subseteq S$, the mapping $S\to \Omega$
marking an element true iff it is in $S'$.}

\Card{Epimorphism}{
    $F$ is {\em epic} or an {\em epi} iff $g_1 f = g_2 f \Rightarrow g_1 = g_2$.

    \Comdiag{
    X \arrow[r,"f"] \&Y \arrow[r, shift left,"g_1"] \arrow[r,"g_2"'] \& Z
    }

    $f$ is injective iff it is epic.
}

\Card{Monomorphism}{
    $F$ is {\em monic} or {\em mono} iff $f g_1 = f g_2 \Rightarrow g_1 = g_2$.

    \Comdiag{
    Z \arrow[r, shift left,"g_1"] \arrow[r,"g_2"'] \&Y \arrow[r,"f"] \& X
    }

    $f$ is surjective iff it is monic.
}

\Card{Isomorphism}{
    A mapping that is epic and monic.
}

\Card[Theorem]{ If $f$ is monomorphic, what can we say about $f'$?
    \Comdiag{
       X\times_Z Y \ar[r, "f'"] \ar[d, "g"'] \& Y \ar[d,"g'"]\\
        X \ar[r, "f"] \& Z
    }
} {
    $f'$ is also monomorphic. Prove via the defn of mononomrpism via $B$ and $X$ (so $c=d$), then the 
    universal property of pullbacks to show $a=b$.
    \Comdiag{
        B \ar[r, shift left, "a"] \ar[r, "b"'] \ar[dr, shift left, "c"] \ar[dr, "d"'] \&
        X\times_Z Y \ar[r, "f'"] \ar[d, "g"'] \& Y \ar[d,"g'"]\\
        \&X \ar[r, "f"] \& Z
    }
}

\Card[Theorem]{
    If $f$ is epic, what can we say about $f'$?
    \Comdiag{
        Z \ar[d, "g"'] \ar[r, "f"] \& Y \ar[d, "g'"] \\
        X \ar[r, "f'"] \& X\sqcup_Z Y
    }
} {$f'$ is also epic. Prove via the defn of epimorpism via $B$ and $Y$ (so $a=b$), then the 
    universal property of pullbacks to show $c=d$.
    \Comdiag{
        Z \ar[r, "f'"] \ar[d, "g"'] \& Y \ar[d,"g'"]\&
            B \ar[l, shift left, "b"] \ar[l, "a"'] \ar[dl, shift left, "d"] \ar[dl, "c"'] \\
        X \ar[r, "f'"] \& X\sqcup_Z Y
    }
}

\Card[Corollary]{
    What is $A$ in
    \Comdiag{A \ar[r, "f'"] \ar[d, "i"] \& \{1\} \ar[d, "True"]\\
        X \ar[r, "f"'] \& \Omega
    }
}{
    $X\times_{\Omega}\{1\}$
}

\Card{Multiset}{
    A sequence $X\equiv (E,B,\pi)$ where $E$ and $B$ are sets and $\pi : E \to B$ is a surjective function. 
}

\Card[Question]{
    Given two multisets $(E,B, \pi)$ and $(E',B', \pi')$, what would a mapping
    between them look like?
}{
    A pair of functions $f_1:E\to E'$ and $f_2:B\to B'$ such that
    \Comdiag{E \ar[r, "f_1"] \ar[d, "\pi"'] \& E'  \ar[d, "\pi'"]\\
        B \ar[r, "f_2"'] \& B'
    }
}


\Card{Pseudomultiset}{
    A multiset where $\pi$ is not surjective.
    A sequence $X\equiv (E,B,\pi)$ where $E$ and $B$ are sets and $\pi : E \to B$. 
}


\Card{Relative set over B}{A pair $(E, \pi)$ with $\pi:E \to R$.}

\Card[Question]{
    Given two relative sets over the same $B$,$(E, \pi)$ and $(E', \pi')$, what would a mapping
    between them look like?
}{
    \Comdiag{E \ar[rr, "f_1"] \ar[dr, "\pi"'] \& \& E'  \ar[dl, "\pi'"]\\
        \& B
    }
}

\Card{$A$-indexed set}{A collection of sets $S_a$, one for each $a\in A$.}


\Card{Closed mapping}{For any element $x\in S$, $f(x)\in S$.}

\Card{Magma}{A set with a closed mapping.}

\Card{Associative mapping}{
    A mapping $f:S\times S$ is associative iff $f(s_1, f(s_2, s_3)) = f(f(s_1, s_2), s_3)$.
}

\Card{Semigroup}{A set with a closed, associative mapping.

A magma whose operation is also associative.}

\card{Abelian group}{
    AKA a commutative group:
    a group whose operation is also commutative: $a\cdot b = b\cdot a$.
}

\Card{Monoid}{
    A set with an operation that is closed and associative,
    and an identity element where $f(x)=x$.

    I.e., a semigroup with an identity.
}

\Card{Trivial monoid}{
    A monoid over a set whose sole element is the identity, $(\{id\}, id, \cdot)$.

    Written as $\underline 1$.
}

\Card{List in a set $X$}{
    A pair $(n, f)$, where $f:\underline n \to X$. May be denoted as
    $(n, f) = [f(1), f(2), \dots, f(n)]$.
}

\Card[notation]{Concatenation of lists, $++$}{
    Given $L=(n, f)$ and $L'=(n',f')$ a new list $L++L' =(n+n', f++f')$ where $f++f'$
    matches $f(x)$ for $x\leq n$, then $f'(x-n)$ for $x>n$.
}

\Card{Free monoid generated by a set $X$}{
    A triplet $($List$(X), [ ], ++)$, where the first element is the set of all lists ($n,
    f)$ where $n=|X|$.
}

\Card{Presented monoid}{
    Let $\sim$ indicate the equivalence relation on List$(G)$ (the set of all lists
    on $G$) generated by $(x\cdot f(i)\cdot y \sim x\cdot f'(i)\cdot y)$, where $f$
    and $f'$ are in List$(G)$.

    The monoid presented by the generator $G$ and relations $f(i), f'(i)$ is the set
    List$(G)\backslash \sim$, the identity $[ ]$, and the operation $++$.
}

\Card{Cyclic monoid}{
    A presented monoid based on a generator with a single element.
}

\Card{Monoid action}{
    Given a monoid $(M, id, \cdot)$ and a set $S$, the action $\circlearrowleft:M\times S
    \to S$ satisfies
    \items{
        \item $id \circlearrowleft s = s, s\in S$
        \item $m \circlearrowleft (n \circlearrowleft s) = (m\cdot n) \circlearrowleft s$
    }
}

\Card{Finite state machine}{
    \items{
    \item A set of symbols, the input alphabet $\Sigma$
    \item A set of states $S$
    \item A transition function $\delta: \Sigma \times S\to S$
    \item An initial state $\in S$
    \item A set of final states $\subseteq S$
    }
}

\Card[Proposition]{How is a finite state machine expressed via monoids?}{
    Given $\delta: \Sigma \times S \to S$, let the monoid be the free monoid based on
    List$(\Sigma)$; then the state machine is an action of the free monoid on $S$.
}

\card{Monoid homomorphism}{
    For $(S, id, \cdot)$ and $(S', id', \cdot')$, a M. H. $f:S\to S'$ has $f(id)=id'$ and
$f(s_1\cdot s_2) = f(s_1)\cdot' f(s_2)$.
}

\card{$\Hom{Mon}(M,M')$}{
    The set of monoid homomorphisms $f:M\to M'$.
}

\card{Trivial monoid homomrphism}{
    Given the trivial monoid $\underline 1$, $\exists$ monoid homomorphisms $f_1:M\to
    \underline 1$ and $f_2: \underline 1 \to M$
}

\card[Proposition]{Given $M=({\mathbb Z}, 0, +)$ and $M'=({\mathbb N}, 0, +)$, what monoid
homomorphisms exist}{
    Only the trivial homomorphism.
}

\card[Notation]{$\Delta_f(\alpha)$}{
    Chain the monoid homomorphism $f:M\to M'$ and the action $\alpha:(S, M') \to S$ to
    generate a new action $\Delta_f(\alpha): (S,M) \to S$. 

    Named {\em the restriction of scalars along $f$}, due to an application to ${\mathbb R}$ and ${\mathbb C}$.
}

\card{Group}{
    A set with a binary operation that is

    \items{
    \item    closed
    \item    associative,
    \item    has an identity, and
    \item    has an inverse
    }

    I.e., a monoid where every element has an inverse.
}

\card{Ring}{
    A set $R$ with two binary operations, annotated $+$ and $\cdot$, where
    \items{
        \item $\{R, +\}$ is an abelian (commutative) group
        \item $\{R, \cdot\}$ is a monoid
        \item The distributive property holds:
            $$a \cdot (b + c) = (a \cdot b) + (a \cdot c)\\
            (b + c) \cdot a = (b \cdot a) + (c \cdot a) $$
    }
}

\card{Poset}{
    Partially ordered set: a set $S$ and relation $\leq$ where
    \items{
        \item $a \leq a$ (reflexive)
        \item $a \leq b$ and $b \leq a \Rightarrow a=b$ (antisymmetry)
        \item $a \leq b$ and $b \leq c \Rightarrow a\leq c$ (transitivity)
    } for all $a, b, c \in S$.

    A preorder with the addition of the antisymmetry condition.
}

\card{Graph}{
    A collection $(V, A, src, tgt)$:
    \items{
    \item $V$ a set (vertices)
    \item $A$ a set (arrows)
    \item $src:A\to V$ a fn giving arrow heads
    \item $tgt:A\to V$ a fn giving arrow tails
    }
}

\card[Exercise]{Express a unidirectional bipartite graph via a commutative diagram with a span}{
    The top triangle $A, L, src$ and bottom triangle $A, R, tgt$ commute:
    \Comdiag{
        \& L \ar[dr,"i_1"]\\
    A \ar[ur, "f"] \ar[dr, "g"] \ar[rr, shift left,"src"] \ar[rr,"tgt"'] \& \& L \sqcup R \\
        \& R \ar[ur,"i_2"]\\
    }
}

\card{Graph homomorphism}{
    Two functions $f_0:V\to V'$ and $f_1:A\to A'$ such that both $src$ and $tgt$ commute:

    \Comdiag{
    A \ar[r,"f_1"] \ar[d, "src"']\& A'  \ar[d, "src'"]  \& \& A \ar[r,"f_1"] \ar[d, "tgt"']\& A'  \ar[d, "tgt'"]  \\
    V \ar[r,"f_0"] \& V' 
    \& \& V \ar[r,"f_0"] \& V' 
    }
}

\card{Discrete graph, $Disc({\bf S})$}{For a set $S$, the graph with:
    \items{
    \item nodes = $S$
    \item arrows = $\emptyset$
    \item sources = trivial function $!: \emptyset \to S$
    \item targets = trivial function $!: \emptyset \to S$
    }
}

\card{Terminal category}{
    The discrete graph for a single element.
}


\card{Binary relation}{For the set $X$, a subset $R\subseteq X\times X$.}

\card[Notation]{$\leq$}{$s\leq s'$ iff $(s, s')\in R$, a binary relation $R\subseteq S\times S$.}

\card{Preorder}{
$\leq$ is a preorder iff
    \items{
    \item $s\leq s$ (reflexivity)
    \item $s\leq s'$ and $s' \leq s'' \Rightarrow s\leq s''$ (transitivity)
    }
}

\card{Partial order}{
    A preorder with antisymmetry.

    $\leq$ is a partial order iff
    \items{
    \item $s\leq s$ (reflexivity)
    \item $s\leq s'$ and $s' \leq s'' \Rightarrow s\leq s''$ (transitivity)
    \item $s\leq s'$ and $s' \leq s \Rightarrow s= s''$ (antisymmetry)
    }
}

\card{Linear order}{
    A partial order with comparability.

    $\leq$ is a linear order iff
    \items{
    \item $s\leq s$ (reflexivity)
    \item $s\leq s'$ and $s' \leq s'' \Rightarrow s\leq s''$ (transitivity)
    \item $s\leq s'$ and $s' \leq s \Rightarrow s= s''$ (antisymmetry)
    \item Either $s\leq s'$ or $s' \leq s$ (comparability)
    }
}

\card{Clique in a preorder}{
    For a preorder $(S, \leq)$, a set $C\in S$ such that for every $s, s' \in C$, $s\leq s'$.
}

\card{Preorder generated by a binary relation}{
    Start with an aribtrary binary relation $R\subseteq S\times S$. 
    Build the preorder generated by $R$, $R''$ via

    \items{
    \item $R'\equiv R \cup \{(s, s)| s\in S\}$ (make it reflexive),
    \item $R''\equiv R' \cup \{(x, z)\in S\times S| (x,y), (y,z)\in S\times S\}$
(make it transitive)
    }
}

\card{For a preorder $(S, \leq)$, the meet of $a, b\in S$}{
    An element $c$ such that
    \items{
    \item $c\leq a$ and $c\leq b$
    \item For any $d\in S$, $d\leq a$ and $d\leq b \Rightarrow d\leq c$
    }

    Note that for the preorder $($set of sets$, \subseteq)$, $c$ is $a\wedge b$.
}

\card{For a preorder $(S, \leq)$, the join of $a, b\in S$}{
    An element $c$ such that
    \items{
    \item $a\leq c$ and $b\leq c$
    \item For any $d\in S$, $a\leq d$ and $b\leq d \Rightarrow c\leq d$
    }

    Note that for the preorder $($set of sets$, \subseteq)$, $c$ is $a\vee b$.
}

\card{Morphism of preorders}{
    Given $(S, \leq)$ and $(S', \leq')$, a function $f:S\to S'$ such that
    $$s_1\leq s_2 \Rightarrow f(s_1)\leq' f(s_2).$$
}

\card{Discrete preorder on $S$}{$$s\leq s' { \rm\ iff\ } s=s'.$$}

\card{Indiscrete preorder on $S$}{$$s\leq s',\  \forall  s, s'\in S.$$}

\card{Category}{
    \items{
    \item A collection of objects, $Ob({\cal C})$
    \item $\forall x, y\in Ob({\cal C})$, the Hom-set $\Hom{\cal C}(x, y)$ listing morphisms $x\to y$
    \item $\forall x\in Ob({\cal C})$, an identity morphism on $x$, $id_x \in \Hom{\cal C}(x, x)$
    \item A composition formula that is associative and correctly handles the identitity morphisms,
 $$\circ: \Hom{\cal C}(y, z) \times \Hom{\cal C}(x, y) \to \Hom{\cal C}(x, z)$$
    }
}

\card[Sub-defintion]{Properties of $\circ$ for a category}{
    \items{
    \item For any $f:x\to y$, $f\circ id_x = f$ and $id_y \circ f = f$
    \item Associativity: with $f:w\to x$, $g:x\to y$, $h:y\to z$, $$h\circ(g\circ f) = (h
\circ g)\circ f$$
    }

}

\card{The category {\bf Set}}{
    \items{
    \item The set of sets $S$
    \item $\forall$ sets $X, Y\in S$, the set of mappings $f:X\to Y$,
    \item including the identity morphism $id_x : X\to X$ for any $X\in S$.
    \item The usual function composition formula
    }
}

\card{The category {\bf Fin}}{
    The category of sets with finite cardinality.
    \items{
    \item Objects are finite sets
    \item Morphisms and composition formula are as with {\bf set}
    }
}

\card{The category {\bf Mon}}{
    \items{
    \item The set of monoids $(M, id, \cdot)$
    \item For any two monoids, the set of monoid homomorphisms between them.
    \item Monoid homomorphisms are defined via functions $f:M\to M'$ and $g:M'\to M''$;
        composing the functions to $g\circ f$ generates a morphism composition.
    }
}

\card{The category {\bf Grp}}{
    \items{
    \item The set of groups $(M, id, \cdot)$ (i.e., monoids where $\cdot$ is invertible)
    \item Morphisms and composition are defined as with {\bf Mon}
    }
}

\card{The category {\bf Grpd}}{
    Groupoids: a category where every morphism is an isomorphism.
}

\card{The category {\bf PrO}}{
    The set of preorders $(S, \leq)$, with the usual morphisms on preorders of the form
$f:S\to S'$. Composition of functions induces composition of $\Hom{Preorders}$
}

\card{The category {\bf FLin}}{The set of finite linear orders $(S, \leq)$.}

\card{The category {\bf Grph}}{The set of graphs and graph homomorphisms between them.}

\card{The category {\bf GrIn}}{
    The graph indexing category
    \items{
    \item Objects: \{A, V\} (one arrow, one vertex)
    \item Mappings: $src: A\to V$; $tgt: A\to V$
    }
}

\card{Free category generated by a graph}{
    Vertices are category objects; arrows are morphisms. Identity arrows and all arrow
    compositions are included.
}

\card[Notation]{$[1] \in {\bf Cat}$}{
    The free category generated by the graph \Comdiag{v_1 \ar[r] \& v_2}
}

\card{Isomorphism between objects in a category}{
    A morphism $f:X\to Y$, $X,Y\in Ob({\cal C})$, such that $\exists G:Y\to X$ such that
    $g\circ f = id_x$ and $f\circ g = id_y$.
}

\card{Endomorphism}{A mapping from an object to itself.}

\card{Automorphism}{
    An isomorphic mapping from an object to itself.

    An endomorphism that is also an isomorphism.
}

\card{Functor}{
    For two categories {\bf C} and {\bf D}:
    \items{
    \item Object map: for every $x\in Ob(C)$, an $F(x)\in Ob(D)$
    \item Function map: for every $f:A\to B$, an $F(f)$ where  $F(f:A\to B) = F(f): F(A)\to F(B)$
    \item Preserves identity morphisms: $F(id_x)=id_{f(x)}$
    \item Preserves composition: $F(g\circ f)=F(g)\circ F(f)$
    }
}

\card[Exercise]{Express monoid actions via categories}{
    Define {\bf Mon} with one object, morphisms representing the monoid's elements, and composition
    in {\bf Mon} representing the monoid's function.

    Define {\bf Set} as one object, and one element of $Hom(S,S)$ per $S\to S$ morphism.

    Curry $\circlearrowleft:M\times S \to S$ to $\circlearrowleft:M\to Hom(S, S)$, giving the
    functor {\bf Mon} $\to$ {\bf Set}.
}

\card{Natural transformation}{
    Given two categories {\bf C} and {\bf D}, and two functors $F$ and $G$ mapping from
    {\bf C} to {\bf D}. Then $\alpha$ is a natural transformation from $F$ to $G$ iff, for any
    two elements $a, b \in {\bf C}$, this commutes:
    \Comdiag{F(a)  \ar[r, "\alpha_a"] \ar[d, "F(h)"] \& F(a)  \ar[d, "G(h)"]\\
        G(a) \ar[r, "\alpha_b"] \& G(b)
    }
}

\card{The category {\bf Fun(C, D)}}{
    For categories {\bf C} and {\bf D}, objects in the category {\bf Fun(C, D)}
    are the set of functors mapping {\bf C} to {\bf D}, and arrows are the natural
    transformations between those functors.
}

\end{document}
